---
title: "Lecture 31"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Data: flipped classrooms?

Data set has 375 rows (one per student), and the following variables:

* `professor`: which professor taught the class (1 -- 15)
* `style`: which teaching style the professor used (no flip, some flip, fully flipped)
* `score`: the student's score on the final exam


## Visualizing the data

```{r echo=F, message=F, warning=F, fig.width=9, fig.height=6, fig.align='center'}

library(tidyverse)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

diffs <- mean_scores - rep(c(80, 87, 83), each=5)

mean_scores <- rep(c(80, 87, 83), each=5) + 0.5*diffs

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20))
```

## Mixed effects model

**Linear mixed effects model:** Let $Score_{ij}$ be the score of student $j$ in class $i$

\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$


## Fitting mixed effects models

```{r eval=F}
library(lme4)
m1 <- lmer(score ~ style + (1|professor), 
           data = teaching)
summary(m1)
```


## Fitting mixed effects models

```{r, include=F}
set.seed(3)
mean_scores <- c(rnorm(5, 80, 6), rnorm(5, 87, 6), rnorm(5, 83, 6))

teaching <- data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = as.factor(professor))
```

```{r message=F}
library(lme4)
```

```{r, output.lines = 11:14}
m1 <- lmer(score ~ style + (1|professor), 
           data = teaching)
summary(m1)
```


## Fitting mixed effects models

```{r, output.lines = 17:21}
m1 <- lmer(score ~ style + (1|professor), 
           data = teaching)
summary(m1)
```


## Intra-class correlation

:::: {.columns}

::: {.column width="50%"}
```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4, fig.align='center'}

library(latex2exp)

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_3 <- data.frame(score = 80, professor = 5.5,
                         style = factor("full flip",
                                        levels = c("no flip", "some flip", "full flip")))

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 6),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  scale_y_continuous(breaks = c(80, 83, 87),
                     labels = c(TeX("$\\beta_0"),
                                TeX("$\\beta_0 + \\beta_2"),
                                TeX("$\\beta_0 + \\beta_1"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20)) +
  geom_segment(aes(x = 0.5, y = 80,
                   xend = 5.5, yend = 80),
               data = ann_text_1,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 87,
                   xend = 5.5, yend = 87),
               data = ann_text_2,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 83,
                   xend = 5.5, yend = 83),
               data = ann_text_3,
               lwd = 1, lty=3)
```

$\sigma_\varepsilon^2$ is large relative to $\sigma_u^2$
:::

::: {.column width="50%"}
```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4, fig.align='center'}

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_3 <- data.frame(score = 80, professor = 5.5,
                         style = factor("full flip",
                                        levels = c("no flip", "some flip", "full flip")))

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

diffs <- mean_scores - rep(c(80, 87, 83), each=5)

mean_scores <- rep(c(80, 87, 83), each=5) + 3*diffs

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  scale_y_continuous(breaks = c(80, 83, 87),
                     labels = c(TeX("$\\beta_0"),
                                TeX("$\\beta_0 + \\beta_2"),
                                TeX("$\\beta_0 + \\beta_1"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20)) +
  geom_segment(aes(x = 0.5, y = 80,
                   xend = 5.5, yend = 80),
               data = ann_text_1,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 87,
                   xend = 5.5, yend = 87),
               data = ann_text_2,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 83,
                   xend = 5.5, yend = 83),
               data = ann_text_3,
               lwd = 1, lty=3)
```

$\sigma_\varepsilon^2$ is small relative to $\sigma_u^2$
:::

::::

**Intra-class correlation:** $\rho_{group} = \dfrac{\sigma_u^2}{\sigma_u^2 + \sigma_\varepsilon^2} = \dfrac{\text{between group variance}}{\text{total variance}}$


## Intra-class correlation

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$

* $\widehat{\beta}_0 = 77.66$, $\hspace{1cm} \widehat{\beta}_1 = 11.07$, $\hspace{1cm} \widehat{\beta}_2 = 2.81$
* $\widehat{\sigma}_\varepsilon^2 = 4.25$, $\hspace{1cm} \widehat{\sigma}_u^2 = 21.37$

$$\widehat{\rho}_{group} = \dfrac{21.37}{21.37 + 4.25} = 0.83$$

So 83% of the variation in student's scores can be explained by differences in average scores from class to class (after accounting for teaching style). That's huge!


## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_32.html](https://sta712-f23.github.io/class_activities/ca_lecture_32.html)

