---
title: "Lecture 19"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(knitr)
library(gridExtra)
library(statmod)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

## New data

Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: number of drinks the student reports consuming
* `sex`: whether the student identifies as male
* `OffCampus`: whether the student lives off campus
* `FirstYear`: whether the student is a first-year student

Our goal: model the number of drinks students report consuming.

## EDA: drinks

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)

wdrinks <- read.csv("~/Documents/Teaching/sta279-s22.github.io/slides/weekendDrinks.csv")

FirstYear <- ifelse( wdrinks$dorm %in% c("mohn","kittlesby", "kildahl"), "TRUE", "FALSE")
OffCampus <- ifelse( wdrinks$dorm == "off campus", "TRUE", "FALSE")

wdrinks <- cbind(wdrinks, FirstYear, OffCampus)

wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25))
```

What do you notice about this distribution?

## Comparisons with Poisson distributions

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=7}
library(gridExtra)
set.seed(1)
p1 <- wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Observed data", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

p2 <- data.frame(x = rpois(100, 1)) %>%
  ggplot(aes(x = x)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Poisson(1)", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

p3 <- data.frame(x = rpois(100, 2)) %>%
  ggplot(aes(x = x)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Poisson(2)", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

p4 <- data.frame(x = rpois(100, 5)) %>%
  ggplot(aes(x = x)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(title = "Poisson(5)", x = "", y = "") +
  scale_x_continuous(limits = c(-1, 25)) +
  scale_y_continuous(limits = c(0, 45))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

## Excess zeros

*Why might there be excess 0s in the data, and why is that a problem for modeling the number of drinks consumed?*

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
wdrinks %>%
  ggplot(aes(x = drinks)) +
  geom_bar() +
  theme_bw() +
  theme(text = element_text(size = 25))
```

## Hurdle models: model the zeros separately

## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_19.html](https://sta712-f23.github.io/class_activities/ca_lecture_19.html)

