---
title: "Lecture 15"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Motivating example: air pollution data

* Data on Chicago air quality and death between 1987 and 2000
* Variables include:
  - deaths
  - ozone concentration
  - sulphur dioxide concentration
  - temperature

## Motivating example: air pollution data

$$Deaths_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Ozone_i$$

## Quantile residual plot

```{r, include=F}
library(statmod)
library(gamair)
library(tidyverse)

data("chicago")

m1 <- glm(death ~ o3median, data = chicago, family = poisson)
```

```{r, echo=F, message=F, warning=F}
chicago |>
  drop_na(death, o3median) |>
  ggplot(aes(x = o3median, y = qresid(m1))) +
  geom_point() +
  geom_smooth() +
  theme_bw() + 
  labs(x = "Ozone", y = "Quantile residuals")
```

## GOF test

```{r}
m1$deviance

m1$df.residual

pchisq(m1$deviance, m1$df.residual, lower.tail=F)
```

## Overdisperion

**Overdispersion** occurs when the response variable $Y_i$ has greater variability than the model accounts for

## Recap: sandwich estimator for GLMs

## Assumptions about both mean and variance

## Quasi-Poisson models

## Example: Chicago air quality

Poisson model:

```{r, echo=F}
m1 <- glm(death ~ o3median, data = chicago, family = poisson)
summary(m1)$coefficients
```

Quasi-Poisson model:

```{r, echo=F}
m2 <- glm(death ~ o3median, data = chicago, family = quasipoisson)
summary(m2)
```


## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_15.html](https://sta712-f23.github.io/class_activities/ca_lecture_15.html)

  
