---
title: "Lecture 11"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Recap: estimating $\phi$

## Recap: Inference when $\phi$ is known

## Inference when $\phi$ is unknown

## Data

2015 Family Income and Expenditure Survey (FIES) on households in the Phillipines. Variables include

* `age`: age of the head of household
* `numLT5`: number in the household under 5 years old
* `total`: total number of people other than head of household
* `roof`: type of roof (stronger material can sometimes be used as a proxy for greater wealth)
* `location`: where the house is located (Central Luzon, Davao Region, Ilocos Region, Metro Manila, or Visayas)


## Poisson regression model

$Y_i =$ number of people in household other than head

$$Y_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$

## Model assumptions

$Y_i =$ number of people in household other than head

$$Y_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$

**Question:** What assumptions does this Poisson regression model make?

## Model assumptions

$Y_i =$ number of people in household other than head

$$Y_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 Age_i$$

* **Shape:** The shape of the regression model is correct 
* **Independence:** The observations are independent
* **Poisson distribution:** A Poisson distribution is a good choice for $Y_i$

**Question:** How could I assess these assumptions?

## Shape: log empirical means plot

```{r include=F}
library(tidyverse)
fies <- read_csv("~/Documents/Teaching/sta214-f22.github.io/slides/fies.csv")

fies %>%
  ggplot(aes(x = `Total Number of Family members` - 1)) +
  geom_bar() +
  labs(x = "Number of family members") +
  theme_bw() +
  theme(text = element_text(size = 25))

fies <- fies %>%
  mutate(total = `Total Number of Family members`,
         age = `Household Head Age`)
```

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}
set.seed(6)
fies %>%
  group_by(age) %>%
  summarize(emp_mean = log(mean(total)),
            age_mean = mean(age)) %>%
  ungroup() %>%
  ggplot(aes(x = age_mean, y = emp_mean)) +
  geom_jitter(size = 2, width=0.3, height=0.3) +
  labs(x = "Age", y = "log(Average number of people)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

## Shape: log empirical means plot

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}
set.seed(6)
fies %>%
  group_by(age) %>%
  summarize(emp_mean = log(mean(total)),
            age_mean = mean(age)) %>%
  ungroup() %>%
  ggplot(aes(x = age_mean, y = emp_mean)) +
  geom_jitter(size = 2, width=0.3, height=0.3) +
  labs(x = "Age", y = "log(Average number of people)") +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  geom_smooth(se=F, method="lm", formula = y ~ poly(x,2), lwd=1.2)
```

## Shape: quantile residual plot

```{r, eval=F}
m1 <- glm(total ~ age, 
          data = fies, family = poisson)
m2 <- glm(total ~ poly(age, 2), 
          data = fies, family = poisson)
```

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=10, fig.height=4}
library(statmod)
library(gridExtra)
m1 <- glm(total ~ age, 
          data = fies, family = poisson)
m2 <- glm(total ~ poly(age, 2), 
          data = fies, family = poisson)
p1 <- fies %>%
  mutate(resids = qresid(m1)) %>%
  ggplot(aes(x = age, y = resids)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals", 
       title = "No transformation on Age")
p2 <- fies %>%
  mutate(resids = qresid(m2)) %>%
  ggplot(aes(x = age, y = resids)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals", 
       title = "Second order polynomial")
grid.arrange(p1, p2, ncol=2)
```

## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_11.html](https://sta712-f23.github.io/class_activities/ca_lecture_11.html)


## Class activity

```{r, echo=F, message=FALSE, warning=F, fig.align='center', fig.width=10, fig.height=7}
set.seed(1)
r <- 0.5
n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x))
y2 <- rpois(n, lambda = exp(x^2))
y3 <- rnbinom(n, size=r, mu=exp(x))
y4 <- rnbinom(n, size=r, mu=exp(x^2))

m1 <- glm(y1 ~ x, family = poisson)
m2 <- glm(y2 ~ x, family = poisson)
m3 <- glm(y3 ~ x, family = poisson)
m4 <- glm(y4 ~ x, family = poisson)

p1 <- data.frame(x = x, resids = qresid(m1)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals", 
       title = "Poisson data, shape assumption satisfied") +
  scale_y_continuous(limits=c(-3, 6))
p2 <- data.frame(x = x, resids = qresid(m2)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals", 
       title = "Poisson data, shape assumption violated") +
  scale_y_continuous(limits=c(-3, 6))
p3 <- data.frame(x = x, resids = qresid(m3)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals", 
       title = "Non-Poisson data, shape assumption satisfied") +
  scale_y_continuous(limits=c(-3, 6))
p4 <- data.frame(x = x, resids = qresid(m4)) %>%
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "Age", y = "Quantile residuals", 
       title = "Non-Poisson data, shape assumption violated") +
  scale_y_continuous(limits=c(-3, 6))

grid.arrange(p1, p2, p3, p4, ncol=2)
```

## Using quantile residual plots

We can use the quantile residual plot to assess the shape and distribution assumptions:
