---
title: "Lecture 35"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Data: flipped classrooms?

Data set has 375 rows (one per student), and the following variables:

* `professor`: which professor taught the class (1 -- 15)
* `style`: which teaching style the professor used (no flip, some flip, fully flipped)
* `score`: the student's score on the final exam


## Inference with linear models

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_{i}
\end{align}

**Research question:** Is there a relationship between teaching style and student score?

What are my null and alternative hypotheses, in terms of one or more model parameters?


## Inference with linear models

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_{i}
\end{align}

**Research question:** Is there a relationship between teaching style and student score?

$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$

What test would I use to test these hypotheses?

## F tests

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_{i}
\end{align}

**Research question:** Is there a relationship between teaching style and student score?

$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$

What are my degrees of freedom for the F test?

## F tests for mixed effects models

\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

**Research question:** Is there a relationship between teaching style and student score?

What are my null and alternative hypotheses, in terms of one or more model parameters?

## F tests for mixed effects models

\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

**Research question:** Is there a relationship between teaching style and student score?

$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$

**Test:** We will use an F test again

* numerator df = number of parameters tested = 2
* denominator df = ??

## What *are* degrees of freedom?


## Denominator degrees of freedom for mixed models

\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

$H_0: \beta_1 = \beta_2 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$

**Test:** We will use an F test again

* numerator df = number of parameters tested = 2
* denominator df = 

$$\text{number of independent observations} - \text{number of parameters}$$

Are all observations independent?


## Denominator degrees of freedom for mixed models


## Approximating the degrees of freedom

```{r, include=F}
library(lme4)
library(lmerTest)
```


```{r}
groups <- rep(1:30, each=10)
sigma_u <- 0.1
sigma_e <- 0.5

u <- rnorm(30, sd=sigma_u)
x1 <- rnorm(300)
y <- 1 + u[groups] + 0.5*x1 + rnorm(300, sd=sigma_e)

m1 <- lmer(y ~ x1 + (1|groups))
summary(m1)$coefficients
```

## Approximating the degrees of freedom

```{r}
groups <- rep(1:30, each=10)
sigma_u <- 1
sigma_e <- 0.5

u <- rnorm(30, sd=sigma_u)
x1 <- rnorm(300)
y <- 1 + u[groups] + 0.5*x1 + rnorm(300, sd=sigma_e)

m1 <- lmer(y ~ x1 + (1|groups))
summary(m1)$coefficients
```


## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_35.html](https://sta712-f23.github.io/class_activities/ca_lecture_35.html)

