---
title: "Lecture 20"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(knitr)
library(gridExtra)
library(statmod)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

```

## Data from last time

Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: number of drinks the student reports consuming
* `sex`: whether the student identifies as male
* `OffCampus`: whether the student lives off campus
* `FirstYear`: whether the student is a first-year student

Our goal: model the number of drinks students report consuming.

```{r include=F}
library(tidyverse)

wdrinks <- read.csv("~/Documents/Teaching/sta279-s22.github.io/slides/weekendDrinks.csv")

FirstYear <- ifelse( wdrinks$dorm %in% c("mohn","kittlesby", "kildahl"), "TRUE", "FALSE")
OffCampus <- ifelse( wdrinks$dorm == "off campus", "TRUE", "FALSE")

wdrinks <- cbind(wdrinks, FirstYear, OffCampus)
```

## Recap: Poisson hurdle model

## Fitting Poisson hurdle models

## Fitting the model in R

```{r, message=F, warning=F}
library(pscl)

m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m1$coefficients
```

## Model assumptions

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m1$coefficients
```

What assumptions does this model make?

## Model diagnostics

## Rootograms

```{r, message=F, warning=F}
library(countreg)
rootogram(m1)
```

## Other examples with rootograms

```{r, fig.align='center', fig.width=12, fig.height=5}
par(mfrow=c(1, 2))

x <- rnorm(1000)
y <- rnbinom(1000, 0.5, mu=exp(1.5 + 0.2*x))
m1 <- glm(y ~ x, family = poisson)
m2 <- glm.nb(y ~ x)

rootogram(m1)
rootogram(m2)
```

```{r, include=F}
par(mfrow=c(1,1))
```



## Hurdle models for count data

$$P(Y_i > 0) = p_i \hspace{1cm} g_{zero}(p_i) = \gamma^T X_i$$

$$Y_i | (Y_i > 0) \sim ZT(\lambda_i) \hspace{1cm} g_{count}(\lambda_i) = \beta^T X_i$$

## Negative binomial hurdle model

```{r}
m2 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "negbin", zero.dist = "binomial",
             data = wdrinks)

rootogram(m2)
```


## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_20.html](https://sta712-f23.github.io/class_activities/ca_lecture_20.html)

