---
title: "Lecture 21"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Last time

Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: number of drinks the student reports consuming
* `sex`: whether the student identifies as male
* `OffCampus`: whether the student lives off campus
* `FirstYear`: whether the student is a first-year student

Our goal: model the number of drinks students report consuming.

## Last time

```{r include=F}
library(tidyverse)

wdrinks <- read.csv("~/Documents/Teaching/sta279-s22.github.io/slides/weekendDrinks.csv")

FirstYear <- ifelse( wdrinks$dorm %in% c("mohn","kittlesby", "kildahl"), "TRUE", "FALSE")
OffCampus <- ifelse( wdrinks$dorm == "off campus", "TRUE", "FALSE")

wdrinks <- cbind(wdrinks, FirstYear, OffCampus)
```

```{r, message=F, warning=F}
library(pscl)

m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m1$coefficients
```

**Question:** I want to know whether there is a relationship between sex and the number of drinks a student reports consuming (after accounting for other variables). What hypotheses should I test?

## Hypothesis tests

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)
m2 <- hurdle(drinks ~ FirstYear + OffCampus,
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

2*(m1$loglik - m2$loglik)
pchisq(2*(m1$loglik - m2$loglik), 2, lower.tail=F)
```

## Hypothesis tests

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m1$coefficients
```

**Question:** I want to know whether there is a relationship between sex and whether a student reports consuming *any* drinks. What hypotheses should I test?

## Hypothesis tests

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m2 <- hurdle(drinks ~ sex + FirstYear + OffCampus | FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)
m2$coefficients
```

## Hypothesis tests

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m2 <- hurdle(drinks ~ sex + FirstYear + OffCampus | FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

pchisq(2*(m1$loglik - m2$loglik), df=1, lower.tail=F)
```

## Hypothesis tests

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

m1$coefficients
```

**Question:** *Among students who report at least one drink*, I want to know whether male students tend to drink *more*. What hypotheses should I test?

## Hypothesis tests

```{r}
m1 <- hurdle(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson", zero.dist = "binomial",
             data = wdrinks)

summary(m1)$coefficients
pnorm(5.233, lower.tail=F)
```

## What's next

* Problem: excess zeros!
* Solution so far: hurdle model (Poisson, negative binomial, etc.)
* Alternative method: zero-inflated models

## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_21.html](https://sta712-f23.github.io/class_activities/ca_lecture_21.html)


