---
title: "Lecture 31"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Data: flipped classrooms?

* A *flipped classroom* involves students watching lectures at home, and doing activities during class time
* There is debate about the pros and cons of this teaching method
* Here we will look at simulated data from an experiment with flipped classrooms


## Data: flipped classrooms?

* 15 classes of introductory statistics
* 25 students in each class (so 375 students total)
* Each class taught by a different professor
* Each professor randomly assigned a teaching style: No flip, Some flip, and Fully flipped
* At the end of the semester, we give all the students in all the classes the same exam, and compare their results


## Data: flipped classrooms?

Data set has 375 rows (one per student), and the following variables:

* `professor`: which professor taught the class (1 -- 15)
* `style`: which teaching style the professor used (no flip, some flip, fully flipped)
* `score`: the student's score on the final exam


## Considering results

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=5, fig.align='center'}

library(tidyverse)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

diffs <- mean_scores - rep(c(80, 87, 83), each=5)

mean_scores <- rep(c(80, 87, 83), each=5) + 0.5*diffs

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20))
```

What do you notice about the scores?


## Considering results

```{r echo=F, message=F, warning=F, fig.width=8, fig.height=5, fig.align='center'}

library(tidyverse)

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

diffs <- mean_scores - rep(c(80, 87, 83), each=5)

mean_scores <- rep(c(80, 87, 83), each=5) + 0.5*diffs

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20))
```

* There may be some differences between styles
* There may be some differences between professors

## Considering results

Suppose we notice that, on average, students in the “Some Flipped” classes have higher scores than students in the “Fully Flipped” classes. What might explain this difference?


## Considering results

Suppose we notice that, on average, students in the “Some Flipped” classes have higher scores than students in the “Fully Flipped” classes. What might explain this difference?

* The "Some Flipped" method may lead to higher test results.
* The professors assigned to teach "Some Flipped" may teach in such a way that their scores are higher than those in the "Fully Flipped" group (more experience, etc.).
* The students in the "Some Flipped" classes may have been stronger than those in the "Fully Flipped" group.


## Different effects

* *Effect of interest* (*treatment effect*): The "Some Flipped" method may lead to higher test results; *the treatment imposed by the researchers has an effect on the outcome.*

* *Group effect*: The professors assigned to teach "Some Flipped" may have had an impact on the test scores; *the group the students are in has an effect on the outcome.*

* *Individual effect*: The students in the "Some Flipped" classes may have been stronger than those in the "Fully Flipped" group; *the individuals’ characteristics or abilities have an effect on the outcome.*


## Writing down a model

*Score* is a continuous response, so we can go back to linear models:

$$Score_i = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_i$$
$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

Which effects does this model capture?


## Assumptions

$$Score_i = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_i$$
$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

What does this model assume about group effects (differences between professors)?


## Assumptions

$$Score_i = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_i$$
$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

What does this model assume about correlation within a class?


## Writing down a model

$$Score_i = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_i$$
$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

How can I incorporate systematic differences between classes?

## Writing down a model

$$Score_i = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \varepsilon_i$$
$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

Add a variable for the different professors:

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \\
& \hspace{1cm} \beta_3 \text{Class2}_i + \cdots + \beta_{16} \text{Class15}_i + \varepsilon_i
\end{align}

$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$


## Writing down a model

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \\
& \hspace{1cm} \beta_3 \text{Class2}_i + \cdots + \beta_{16} \text{Class15}_i + \varepsilon_i
\end{align}

$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

How many parameters did we add to the model to capture class differences?


## Writing down a model

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \\
& \hspace{1cm} \beta_3 \text{Class2}_i + \cdots + \beta_{16} \text{Class15}_i + \varepsilon_i
\end{align}

$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

Do we want to do inference on $\beta_3$,..., $\beta_{16}$ ?

## Our first mixed effects model

**Linear model:**

\begin{align}
Score_i &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + \\
& \hspace{1cm} \beta_3 \text{Class2}_i + \cdots + \beta_{16} \text{Class15}_i + \varepsilon_i
\end{align}

$\varepsilon_i \overset{iid}{\sim} N(0, \sigma_\varepsilon^2)$

**Linear mixed effects model:** Let $Score_{ij}$ be the score of student $j$ in class $i$

$$Score_{ij} = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}$$

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$

## Anatomy of the mixed effects model

**Linear mixed effects model:** Let $Score_{ij}$ be the score of student $j$ in class $i$

$$Score_{ij} = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}$$

$\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma_\varepsilon^2) \hspace{1cm} u_i \overset{iid}{\sim} N(0, \sigma_u^2)$

## Anatomy of the mixed effects model

$$Score_{ij} = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}$$

**Part 1:** Fixed effects (treatment effects)


```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4.5, fig.align='center'}
library(latex2exp)
library(tidyverse)

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

data.frame(score = c(rep(80, 5), rep(87, 5), rep(83, 5)),
           professor = rep(1:5, 3),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=5),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  ggplot() +
  facet_wrap(~style) +
  geom_segment(aes(x = professor - 0.3, y = score,
                   xend = professor + 0.3, yend = score),
               lwd = 2) +
  scale_y_continuous(limits=c(75, 92),
                     breaks = c(80),
                     labels = c(TeX("$\\beta_0"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=30)) +
  geom_segment(aes(x = 5.5, y = 80,
                   xend = 5.5, yend = 87),
               arrow = arrow(length = unit(0.3, "cm")),
               data = ann_text_1,
               lwd = 1) +
  geom_text(x = 4.5, y = 83, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 10, data = ann_text_1) +
  geom_segment(aes(x = 5.5, y = 80,
                   xend = 5.5, yend = 83),
               arrow = arrow(length = unit(0.3, "cm")),
               data = ann_text_2,
               lwd = 1) +
  geom_text(x = 4.5, y = 82, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 10, data = ann_text_2)
```

## Anatomy of the mixed effects model

$$Score_{ij} = \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}$$

**Part 2:** Random effects (group effects)


```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4.5, fig.align='center'}

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_3 <- data.frame(score = 80, professor = 5.5,
                         style = factor("full flip",
                                        levels = c("no flip", "some flip", "full flip")))

data.frame(score = c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85),
           professor = rep(1:5, 3),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=5),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  ggplot() +
  facet_wrap(~style) +
  geom_segment(aes(x = professor - 0.3, y = score,
                   xend = professor + 0.3, yend = score),
               lwd = 2) +
  scale_y_continuous(limits=c(75, 92),
                     breaks = c(80, 83, 87),
                     labels = c(TeX("$\\beta_0"),
                                TeX("$\\beta_0 + \\beta_2"),
                                TeX("$\\beta_0 + \\beta_1"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20)) +
  geom_segment(aes(x = 0.5, y = 80,
                   xend = 5.5, yend = 80),
               data = ann_text_1,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 87,
                   xend = 5.5, yend = 87),
               data = ann_text_2,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 83,
                   xend = 5.5, yend = 83),
               data = ann_text_3,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 1, y = 80,
                   xend = 1, yend = 83.5),
               arrow = arrow(length = unit(0.3, "cm")),
               data = ann_text_1,
               lwd = 1) +
  geom_text(x = 2, y = 82, 
           label = TeX("$u_1", output='character'), 
           parse=TRUE, size = 10, data = ann_text_1) +
  geom_segment(aes(x = 2, y = 87,
                   xend = 2, yend = 80.5),
               arrow = arrow(length = unit(0.3, "cm")),
               data = ann_text_2,
               lwd = 1) +
  geom_text(x = 3, y = 84, 
           label = TeX("$u_7", output='character'), 
           parse=TRUE, size = 10, data = ann_text_2)
```

### Anatomy of the mixed effects model

\begin{align}
Score_{ij} &= \beta_0 + \beta_1 \text{SomeFlipped}_i + \beta_2 \text{FullyFlipped}_i + u_i + \varepsilon_{ij}
\end{align}

**Part 3:** Noise (individual effects)


```{r echo=F, message=F, warning=F, fig.width=8, fig.height=4.5, fig.align='center'}

ann_text_1 <- data.frame(score = 80, professor = 5.5,
                         style = factor("no flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_2 <- data.frame(score = 80, professor = 5.5,
                         style = factor("some flip",
                                        levels = c("no flip", "some flip", "full flip")))

ann_text_3 <- data.frame(score = 80, professor = 5.5,
                         style = factor("full flip",
                                        levels = c("no flip", "some flip", "full flip")))

set.seed(3)

mean_scores <- c(84, 78, 76, 82, 83,
                     82, 80, 90, 91, 87,
                     79, 80, 83, 82, 85)

data.frame(professor = rep(1:15, each=25),
           style = factor(rep(c("no flip", "some flip", "full flip"),
                                 each=125),
                             levels = c("no flip", "some flip", "full flip"))) %>%
  mutate(score = rnorm(375, mean = mean_scores[professor], sd = 2),
         professor = rep(rep(1:5, each=25), 3)) %>%
  ggplot(aes(x = as.factor(professor), y = score)) +
  geom_boxplot(lwd=1) +
  facet_wrap(~style) +
  scale_y_continuous(breaks = c(80, 83, 87),
                     labels = c(TeX("$\\beta_0"),
                                TeX("$\\beta_0 + \\beta_2"),
                                TeX("$\\beta_0 + \\beta_1"))) +
  labs(x = "Professor", y = "Score") +
  theme_classic() +
  theme(text = element_text(size = 30),
        axis.text.x=element_blank(),
        axis.text.y = element_text(size=20)) +
  geom_segment(aes(x = 0.5, y = 80,
                   xend = 5.5, yend = 80),
               data = ann_text_1,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 87,
                   xend = 5.5, yend = 87),
               data = ann_text_2,
               lwd = 1, lty=3) +
  geom_segment(aes(x = 0.5, y = 83,
                   xend = 5.5, yend = 83),
               data = ann_text_3,
               lwd = 1, lty=3)
```