---
title: "Lecture 29"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Motivating example: earthquake data

We have data from the 2015 Gorkha earthquake in Nepal. After the earthquake, a large scale survey was conducted to determine the amount of damage the earthquake caused for homes, businesses and other structures. Variables include:

* `Damage`: the amount of damage suffered by the building (none, moderate, severe)
* `age`: the age of the building (in years)
* `condition`: a de-identified variable recording the condition of the land surrounding the building


## Fisher scoring


## Wald tests

```{r, include=F}
library(tidyverse)
library(nnet)
library(gridExtra)

earthquake <- read.csv("~/Documents/Teaching/sta712-f22.github.io/slides/earthquake_small.csv") %>%
  mutate(Damage = factor(Damage, levels = c("none", "moderate", "severe")))

m1 <- multinom(Damage ~ sqrt(age) + condition,
               data = earthquake)

```

```{r, echo=F, message=F, warning=F}
summary(m1)$coefficients
summary(m1)$standard.errors
```

Suppose we want to know whether there is a relationship between age and the odds of moderate vs. no damage, after accounting for surface condition. What hypotheses would we test?


## Wald tests

```{r, echo=F, message=F, warning=F}
summary(m1)$coefficients
summary(m1)$standard.errors
```

Suppose we want to know whether the relationship between age and the odds of moderate vs. no damage is the *same* as the relationship between age and the odds of severe vs. no damage. What hypotheses would we test?

## Wald tests

```{r}
diff <- t(c(0, -1, 0, 0, 0, 1, 0, 0)) %*% 
  c(t(coef(m1)))
std_err <- sqrt(t(c(0, -1, 0, 0, 0, 1, 0, 0)) %*% 
                  vcov(m1) %*% 
                  c(0, -1, 0, 0, 0, 1, 0, 0))
(diff - 0)/std_err

2*pnorm((diff - 0)/std_err, lower.tail = F)
```


## Likelihood ratio tests

```{r, echo=F, message=F, warning=F}
summary(m1)$coefficients
summary(m1)$standard.errors
```

Suppose we want to know whether there is a relationship between surface condition and damage, after accounting for building age. What hypotheses would we test?


## Likelihood ratio tests

```{r, results = F}
m1 <- multinom(Damage ~ sqrt(age) + condition,
               data = earthquake)
m2 <- multinom(Damage ~ sqrt(age),
               data = earthquake)

pchisq(m2$deviance - m1$deviance, df = 4, 
       lower.tail=F)
```

```{r, echo=F}
pchisq(m2$deviance - m1$deviance, df = 4, lower.tail=F)
```

## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_29.html](https://sta712-f23.github.io/class_activities/ca_lecture_29.html)


