---
title: "Lecture 1"
format: 
  revealjs:
    theme: theme.scss
editor: visual
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(tidyverse)
library(statmod)
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")
```

## Logistic regression recap

Recall the dengue data from last semester:

* Data on Vietnamese children admitted to hospital with possible dengue fever
* Variables include:
  * Age
  * White blood cell count (WBC)
  * Platelet count (PLT)
  * Dengue status (0 = no dengue, 1 = dengue)
  
## Logistic regression recap

I want to model dengue status, with Age, WBC, and PLT as explanatory variables.

What does my model look like?

## Logistic regression recap

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Age_i + \beta_2 WBC_i + \beta_3 PLT_i$$

**Question:** How do I interpret a regression coefficient (e.g. $\beta_1$)?

## Logistic regression recap

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Age_i + \beta_2 WBC_i + \beta_3 PLT_i$$

**Question:** What assumptions does this model make?

## Logistic regression recap

Quantile residual plot:

```{r, echo=F}
m1 <- glm(Dengue ~ Age + WBC + PLT, data = dengue, 
          family = binomial)
dengue |>
  mutate(resids = qresid(m1)) |>
  ggplot(aes(x = Age, y = resids)) +
  geom_point() +
  geom_smooth(se=F) +
  labs(x = "Age", y = "Quantile residuals") +
  theme_bw() +
  theme(text = element_text(size = 15))
```


## Logistic regression recap

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Age_i + \beta_2 WBC_i + \beta_3 PLT_i$$

**Question:** I want to know whether there is relationship between Age and Dengue status, after accounting for WBC and PLT. How can I address this question?

## Logistic regression recap

```{r}
m1 <- glm(Dengue ~ Age + WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)
```


## Logistic regression recap

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Age_i + \beta_2 WBC_i + \beta_3 PLT_i$$

**Question:** Suppose now I want to test $H_0: \beta_2 = \beta_3 = 0$. How do I carry out the likelihood ratio test?

## Logistic regression recap

```{r, echo=T}
m1 <- glm(Dengue ~ Age + WBC + PLT, data = dengue, family = binomial)
m1$deviance
```

```{r, echo=T}
m2 <- glm(Dengue ~ Age, data = dengue, family = binomial)
m2$deviance
```

Test statistic =

## Logistic regression recap

```{r, echo=T}
m1 <- glm(Dengue ~ Age + WBC + PLT, data = dengue, family = binomial)
m1$deviance
```

```{r, echo=T}
m2 <- glm(Dengue ~ Age, data = dengue, family = binomial)
m2$deviance
```

Test statistic = $deviance_{reduced} - deviance_{full} = 1071.6$

How do I calculate a p-value?

## Logistic regression recap

```{r, echo=T}
m1 <- glm(Dengue ~ Age + WBC + PLT, data = dengue, family = binomial)
m1$deviance
```

```{r, echo=T}
m2 <- glm(Dengue ~ Age, data = dengue, family = binomial)
m2$deviance
```

Test statistic = $deviance_{reduced} - deviance_{full} = 1071.6$

```{r}
pchisq(1071.6, df=2, lower.tail=F)
```

## Logistic regression recap

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \frac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Age_i + \beta_2 WBC_i + \beta_3 PLT_i$$

**Question:** The researchers are interested in whether their model does a good job identifying patients with dengue. Do our hypothesis tests address that question?


## Rough course plan

* Logistic regression recap
* Prediction and model selection
* Supplementary skills (research papers, SAPs, simulation)
* Poisson regression and EDMs
* Mis-specified models (overdispersion, zero-inflation, etc.)
* Correlated data

## Course components

* Homework assignments (graded on completion)
* Challenge assignments (graded on mastery)
* Data analysis projects (graded on mastery)
* Semester research project (graded on mastery)
  * Group project
  * Involves written report, final presentation, and intermediate check-points
  * Due next Monday: group members and tentative topic
  
## Reading a research paper

Research papers in the sciences and social sciences typically contain:

* Abstract
* Introduction
* Methods
* Results
* Discussion
* Conclusion

## Reading a research paper

* **Abstract:** overview and key points
* **Introduction:** motivation, background, overview of work
* **Methods:** details on study design, data, statistical analysis
* **Results:** summary of results, including figures, tables, p-values, etc.
* **Discussion:** discussion of results in context of research question
* **Conclusion:** short summary of paper and key results; connection to broader research

## Class activity

Reading the original dengue paper:

[https://sta712-f23.github.io/class_activities/ca_1.pdf](https://sta712-f23.github.io/class_activities/ca_1.pdf)

For next class: 

* finish reading the paper and working through the class activity
* we will discuss the paper on Wednesday

