---
title: "Lecture 26"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Data

Data on the number of fish caught by campers in a state park. We have a sample of 250 groups of park guests who visited the state park. For each group, we record:

* `count`: the number of fish caught by the group
* `camper`: whether the group brought a camper van
* `child`: the number of children in the group
* `persons`: the total number of people in the group
* `LOS`: length of stay (in days)

What model is appropriate if the number of fish is our response?

## Some EDA

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(pscl)
library(MASS)

fish <- read.csv("https://sta214-s23.github.io/homework/fish2.csv")
```

```{r, echo=F, message=F, warning=F}
fish %>%
  ggplot(aes(x = count)) +
  geom_histogram() +
  theme_bw() +
  labs(x = "Fish caught")
```

```{r}
mean(fish$count == 0)
```

## Research question

Park rangers at the state park wonder whether groups with many children tend to catch
fewer fish. They ask you to fit a model to investigate their hypothesis, and they want you
to account for the total number of visitors in the group and whether the group brought a
camper van (they suspect that camper vans make noise that scares away the fish).

## Model

$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - p_i) + p_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - p_i) & y > 0 \end{cases}$

where

$\log \left( \dfrac{p_i}{1 - p_i} \right) = \gamma_0 + \gamma_1 Camper_i + \gamma_2 Child_i + \gamma_3 Persons_i$

$\log \left( \lambda_i \right) = \beta_0 + \beta_1 Camper_i + \beta_2 Child_i + \beta_3 Persons_i$

**Question:** Is there anything else we should add?

## Offsets

$\log \left( \lambda_i \right) = \beta_0 + \beta_1 Camper_i + \beta_2 Child_i + \beta_3 Persons_i + \log(LOS_i)$

## Fitting a model with an offset

```{r}
m1 <- glm(count ~ camper + child + persons, 
          offset = log(LOS), data = fish, family = poisson)
summary(m1)$coefficients
```

## Assessing the Poisson model

```{r}
library(countreg)

rootogram(m1)
```


## ZIP model (with offset)

$P(Y_i = y) = \begin{cases} e^{-\lambda_i}(1 - p_i) + p_i & y = 0 \\ \dfrac{e^{-\lambda_i} \lambda_i^y}{y!}(1 - p_i) & y > 0 \end{cases}$

where

$\log \left( \dfrac{p_i}{1 - p_i} \right) = \gamma_0 + \gamma_1 Camper_i + \gamma_2 Child_i + \gamma_3 Persons_i$

$\log \left( \lambda_i \right) = \beta_0 + \beta_1 Camper_i + \beta_2 Child_i + \beta_3 Persons_i + \log(LOS_i)$

```{r}
m2 <- zeroinfl(count ~ camper + child + persons,
               offset = log(LOS),
               data = fish)
```

## Diagnostics

```{r}
rootogram(m2)
```

## Diagnostics

Quantile residual plot:

```{r, echo=F, message=F, warning=F}
fish |>
  mutate(resids = qresiduals(m2)) |>
  ggplot(aes(x = persons, y = resids)) +
  geom_point() +
  theme_bw() +
  labs(x = "Persons", y = "Quantile residuals") +
  geom_smooth()

```


## Diagnostics

Quantile residual plot:

```{r, echo=F, message=F, warning=F}
fish |>
  mutate(resids = qresiduals(m2)) |>
  ggplot(aes(x = as.factor(persons), y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Persons", y = "Quantile residuals") +
  geom_smooth()

```

## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_26.html](https://sta712-f23.github.io/class_activities/ca_lecture_26.html)
