---
title: "Lecture 7"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(knitr)
library(tidyverse)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```


## Count variables

**Data:** Data on medical facilities and doctors from a sample of 53 different counties in the US. Variables include:

* `MDs`: the number of medical doctors in the county
* `Hospitals`: the number of hospitals in the county

**Research question:** Can we model the relationship between the number of hospitals and the number of doctors?


## Plotting the data

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
library(Stat2Data)
data("CountyHealth")

CountyHealth %>%
  ggplot(aes(x = Hospitals, y = MDs)) +
  geom_point(size = 2) +
  labs(x = "Number of hospitals",
       y = "Number of doctors") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

**Question:** Does a linear regression model seem appropriate for this relationship?


## Trying a transformation

```{r echo=F, message=FALSE, warning=F, fig.align='center', fig.width=7, fig.height=5}

CountyHealth %>%
  ggplot(aes(x = Hospitals, y = sqrt(MDs))) +
  geom_point(size = 2) +
  labs(x = "Number of hospitals",
       y = "sqrt(Number of doctors)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

Is a linear regression model appropriate now?


## Poisson regression


## Fitting the Poisson regression model

```{r, output.lines = 10:18}
m1 <- glm(MDs ~ Hospitals, data = CountyHealth, 
          family = poisson)
summary(m1)
```


## Interpreting the Poisson regression model

```{r, output.lines = 10:12}
m1 <- glm(MDs ~ Hospitals, data = CountyHealth, 
          family = poisson)
summary(m1)
```

## Exponential dispersion models
