---
title: "Lecture 22"
format: 
  revealjs:
    theme: theme.scss
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Last time

Survey data from 77 college students on a dry campus (i.e., alcohol is prohibited) in the US. Survey asks students "How many alcoholic drinks did you consume last weekend?"

* `drinks`: number of drinks the student reports consuming
* `sex`: whether the student identifies as male
* `OffCampus`: whether the student lives off campus
* `FirstYear`: whether the student is a first-year student

**Question:** Why might students report 0 drinks?

## Paper de-brief

How did Lambert (1992) address the problem of excess 0s?

## Zero-inflated Poisson (ZIP) model

## In R

```{r include=F}
library(tidyverse)

wdrinks <- read.csv("~/Documents/Teaching/sta279-s22.github.io/slides/weekendDrinks.csv")

FirstYear <- ifelse( wdrinks$dorm %in% c("mohn","kittlesby", "kildahl"), "TRUE", "FALSE")
OffCampus <- ifelse( wdrinks$dorm == "off campus", "TRUE", "FALSE")

wdrinks <- cbind(wdrinks, FirstYear, OffCampus)
```

```{r, message=F, warning=F}
library(pscl)

m1 <- zeroinfl(drinks ~ sex + FirstYear + OffCampus, 
             dist = "poisson",
             data = wdrinks)

m1$coefficients
```

## Paper de-brief

How did Lamber (1992) propose fitting the ZIP model?

## Fitting ZIP models

## Class activity

[https://sta712-f23.github.io/class_activities/ca_lecture_22.html](https://sta712-f23.github.io/class_activities/ca_lecture_22.html)


