---
title: "Class activity solutions"
format: html
editor: source
author: "Ciaran Evans"
---

1. 

```{r, message=F, warning=F}
library(AER)
library(MASS)
library(pscl)
library(tidyverse)

data("NMES1988")
nmes <- NMES1988 |>
  dplyr::select(visits, hospital, health, chronic, gender, school, insurance)

m1 <- glm(visits ~ ., family = poisson, data = nmes)
```

2.

```{r}
sum(dpois(x = 0, lambda = m1$fitted.values))
```


The Poisson model only predicts about 47 0s, when really there are 683. Quite a big difference!

3.

```{r}
m2 <- glm.nb(visits ~ ., data = nmes)
summary(m2)
```

```{r}
sum(dnbinom(0, size=1.2066, mu = m2$fitted.values))
```

The NB model predicts about 608 0s. Much closer to the truth, but perhaps still slightly underestimating.

4.

```{r}
m3 <- hurdle(visits ~ ., dist = "poisson", zero.dist = "binomial",
             data = nmes)
summary(m3)
```

In the hurdle model, $p_i = P(Y_i > 0)$ (i.e., the probability of at least one physician visit). The coefficient on `school` is positive in the logistic component (0.058541), so individuals with more years of education are more likely to visit a physician at least once, holding other variables fixed.

5.

The coefficients of the count model give us information about individuals who visited a physician at least once (i.e., $Y_i|(Y_i > 0)$). In this case, the coefficient on `healthpoor` is positive (0.253521), so among individuals who visited a physician at least once, those with poor health tend to have more visits (holding other variables fixed).

6.

```{r}
X_count <- model.matrix(m3, "count")
X_zero <- model.matrix(m3, "zero")

p_hat <- 1 - predict(m3, type = "prob")[,1]
lambda_hat <- predict(m3, type="count")

solve(t(X_zero) %*% diag(p_hat * (1 - p_hat)) %*% X_zero)
vcov(m3, "zero")

solve(t(X_count) %*% diag(lambda_hat * exp(lambda_hat)*(-lambda_hat + exp(lambda_hat)- 1) 
                    * p_hat /((exp(lambda_hat) - 1)^2)) %*% X_count)

vcov(m3, "count")
```






