---
title: "Class activity solutions"
format: html
editor: source
author: "Ciaran Evans"
---

```{r, message=F, warning=F}
library(AER)
library(tidyverse)
library(pscl)

data("NMES1988")
nmes <- NMES1988 |>
  select(visits, hospital, health, chronic, gender, school, insurance)
m1 <- hurdle(visits ~ ., dist = "negbin", zero.dist = "binomial", data = nmes)
```

1. Perform a LRT which drops `school` completely from the reduced model:

```{r}
m2 <- hurdle(visits ~ hospital + health + chronic +
               gender + insurance,
             dist = "negbin", zero.dist = "binomial", data = nmes)

pchisq(2*(m1$loglik - m2$loglik), 2, lower.tail=F)
```

So it seems like there is a relationship between years of education and the number of visits.

2. We just want to test `insurance` in the zero hurdle component of the model. We run a one-sided test, since we want to know whether individuals with insurance are *more* likely to visit the doctor at least once.

```{r}
summary(m1)$coefficients
pnorm(7.406, lower.tail=F)
```

3. We just want to test `gender` in the count component of the model. We run a one-sided test, since we want to know whether men visit the doctor *less* often.

```{r}
summary(m1)$coefficients
pnorm(-2.108, lower.tail=T)
```







