---
title: "Class Activity"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

# Part I: Overdispersion

In the first part of this class activity, we will work with data on crimes at 81 US colleges and universities. Our data includes the following variables:

* `type`: college (C) or university (U)
* `nv`: the number of crimes for that institution in the given year
* `enroll1000`: the number of enrolled students, in thousands
* `region`: region of the US C = Central, MW = Midwest, NE = Northeast, SE = Southeast, SW = Southwest, and W = West)

We are interested in comparing the crime rates in different US regions. Our model is

$$Crimes_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i $$
Below are the results of fitting the model with Poisson regression, and with the quasi-Poisson modification to account for overdispersion.

```{r include=F}
library(tidyverse)

crimes <- read_csv("~/Documents/Teaching/sta279-s22.github.io/slides/c_data.csv")
```

```{r}
m1 <- glm(nv ~ region, data = crimes, family = poisson)
summary(m1)
```

<br/>

<br/>

```{r}
m2 <- glm(nv ~ region, data = crimes, family = quasipoisson)
summary(m2)
```

<br/>

<br/>

:::{.question}
#### Question 1

What is our estimated dispersion parameter, $\widehat{\phi}$? Do you think there is overdispersion in our data?
:::

:::{.question}
#### Question 2

Confirm that the standard errors for the quasi-Poisson fit are equal to $\sqrt{\widehat{\phi}}$ times the standard errors for the Poisson fit.
:::

:::{.question}
#### Question 3

We are interested in comparing crime rates for western and central schools. Calculate an appropriate confidence interval.
:::


# Part II: Comparing variance estimates for hypothesis testing

So far, we have covered three different variances for the estimated coefficients $\widehat{\beta}$ in a Poisson regression model:

* The basic approach assumes that the Poisson regression model is correct ($Y_i \sim Poisson(\mu_i)$, and $\mu_i = \exp\{\beta^T X_i\}$), so 

$$Var(\widehat{\beta}) \approx I^{-1}(\beta)$$ 

(the inverse Fisher information)

* The Quasi-Poisson approach no longer assumes $Y_i \sim Poisson(\mu_i)$. Rather, we just assume that $\mathbb{E}[Y_i] = \mu_i = \exp\{\beta^T X_i\}$, and $Var(Y_i) = \phi \mu_i$. Then, 

$$Var(\widehat{\beta}) \approx \phi I^{-1}(\beta)$$

* The sandwich estimator approach weakens the Quasi-Poisson assumptions still further. Now we just assume that $\mathbb{E}[Y_i] = \mu_i = \exp\{\beta^T X_i\}$. Then,

$$Var(\widehat{\beta}) \approx J_n^{-1}(\beta) V_n(\beta) J_n^{-1}(\beta)$$

where $J_n(\beta) = -\mathbb{E}[U'(\beta)]$ and $V_n(\beta) = Var(U(\beta))$

Our goal in this class activity is to compare the behavior of these different approaches. We will assume we have correctly specified the mean $\mu_i$ as a function of $X_i$, so $\widehat{\beta}$ is consistent for the true model parameters in all three approaches. However, depending on the distribution of $Y_i$, the variance estimates may or may not be good.

## Hypothesis testing with Poisson data

Suppose we assume that

$$Y_i \sim Poisson(\mu_i) \hspace{1cm} \log(\mu_i) = \beta_0 + \beta_1 X_i$$

Furthermore, we want to test $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$.

The code below estimates type I error control for the three different methods, with a sample size $n = 500$, when the Poisson assumption is **correct**:

```{r, eval=F}
library(sandwich)

n <- 500
nsim <- 1000

pvals_naive <- rep(NA, nsim)
pvals_quasi <- rep(NA, nsim)
pvals_sandwich <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n, mean=0, sd=1)
  y <- rpois(n, lambda = exp(2 + 0*x))
  
  m1 <- glm(y ~ x, family = poisson)
  m2 <- glm(y ~ x, family = quasipoisson)
  cov_mat <- sandwich(m1)
  
  pvals_naive[i] <- summary(m1)$coefficients[2,4]
  pvals_quasi[i] <- summary(m2)$coefficients[2,4]
  
  test_stat_sandwich <- m1$coefficients[2]/sqrt(cov_mat[2,2])
  pvals_sandwich[i] <- 2*pnorm(abs(test_stat_sandwich), lower.tail=F)
}


mean(pvals_naive < 0.05)
mean(pvals_quasi < 0.05)
mean(pvals_sandwich < 0.05)
```

:::{.question}
#### Question 4

Run the code, and verify that all three methods approximately control the type I error rate at $\alpha = 0.05$.
:::

:::{.question}
#### Question 5

Re-run the code, but this time use a sample size $n = 20$. What happens to type I error control for the three methods?
:::

Question 5 helps answer the question: why don't we always just use the sandwich estimator? The sandwich estimator assumes less about the variance of $Y_i$, and consequently has to estimate more. This means that the sandwich estimator isn't as accurate for small sample sizes.

## Hypothesis testing with "Quasi-Poisson" data

In the previous section, you saw how each of our three approaches performed when the data came from a Poisson distribution. Now let’s investigate what happens when the quasi-Poisson assumption (variance is a linear function of the mean) is satisfied.

Technically, there is no "quasi-Poisson distribution", so we will need to create our own function for simulating data which (approximately) satisfies the quasi-Poisson assumption. The code below will generate this for a given dispersion parameter $\phi = 5$:

```{r, eval=F}
# Function for simulating quasi-Poisson (overdispersed Poisson) data
rqpois <- function(n, mean, dispersion){
  return(rnbinom(n, mu = mean, size = mean/(dispersion - 1)))
}

x <- rnorm(n, mean=0, sd=1)
y <- rqpois(n, mean = exp(2 + 0*x), dispersion = 5)
```


:::{.question}
#### Question 6

Re-run the code from above, but this time use a sample size $n = 500$ and generate "quasi-Poisson" data. What happens to type I error control for the three methods?
:::

## Hypothesis testing with negative binomial data

Next, let's simulate $Y_i$ from a Negative Binomial distribution (so the Poisson distribution is violated, and the Quasi-Poisson distribution is also violated because variance is no longer a linear function of the mean).

:::{.question}
#### Question 7

Re-run the code from above, but this time use a sample size $n = 500$ and generate negative binomial data. What happens to type I error control for the three methods?
:::

**Important:** the type I error for the quasi-Poisson method is still good! Why? Because, if $H_0$ is true for this simple model, then $Var(Y_i)$ does not depend on $X_i$. So, we don't really need to capture the mean-variance relationship correctly.

## Confidence intervals with negative binomial data

To see when the quasi-Poisson method fails, suppose we have data

$$Y_i \sim NB(r=0.5, p_i)$$

$$\log(\mu_i) = 1 + X_i$$
where $X_i \sim N(0, 1)$. We want to calculate a 95% confidence interval for $\beta_1$ (here $\beta_1 = 1$).

The following code estimates coverage of 95% confidence intervals for the three methods:

```{r, eval=F}
n <- 500
nsim <- 1000

covers_naive <- rep(NA, nsim)
covers_quasi <- rep(NA, nsim)
covers_sandwich <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n, mean=0, sd=1)
  y <- rnbinom(n, size = 0.5, mu = exp(1 + 1*x))
  
  m1 <- glm(y ~ x, family = poisson)
  m2 <- glm(y ~ x, family = quasipoisson)
  cov_mat <- sandwich(m1)
  
  upper_naive <- m1$coefficients[2] + 1.96 * summary(m1)$coefficients[2, 2]
  lower_naive <- m1$coefficients[2] - 1.96 * summary(m1)$coefficients[2, 2]
  covers_naive[i] <- upper_naive > 1 && lower_naive < 1
  
  upper_quasi <- m2$coefficients[2] + qt(0.975, n-2)*summary(m2)$coefficients[2, 2]
  lower_quasi <- m2$coefficients[2] - qt(0.975, n-2)*summary(m2)$coefficients[2, 2]
  covers_quasi[i] <- upper_quasi > 1 && lower_quasi < 1
  
  upper_sandwich <- m1$coefficients[2] + 1.96 * sqrt(cov_mat[2,2])
  lower_sandwich <- m1$coefficients[2] - 1.96 * sqrt(cov_mat[2,2])
  covers_sandwich[i] <- upper_sandwich > 1 && lower_sandwich < 1
}


mean(covers_naive)
mean(covers_quasi)
mean(covers_sandwich)
```

:::{.question}
#### Question 8

Run the code above. How does coverage perform for the different methods?
:::


## Key take-aways

* When data is Poisson, all three methods work (they all estimate the correct variance)
* However, the sandwich estimator needs larger sample sizes
* When data is not Poisson, the basic Poisson regression approach (variance = inverse Fisher information for Poisson) fails (poor control of type I error, poor coverage of confidence intervals)
* If $Var(Y_i) = \phi \mu_i$ (a linear relationship between mean and variance), then both the quasi-Poisson and sandwich estimators work
* If there is really *not* a linear relationship between mean and variance (e.g., NB data with a quadratic relationship between mean and variance), then the quasi-Poisson approach will do better than the basic Poisson approach, but worse than the sandwich estimator


