---
title: "Class activity solutions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(pscl)

fish <- read.csv("https://sta214-s23.github.io/homework/fish2.csv")
```

1.

```{r}
m1 <- zeroinfl(count ~ camper + child + persons,
               offset = log(LOS),
               data = fish)
summary(m1)
```

2.

An increase of one in the number of children in the group is associated with an increase in the odds that the group did not fish by a factor of $\exp\{2.08\} = 8$, and a decrease in the number of fish caught *per day*, for groups which *did* fish, by a factor of $\exp\{-0.56\} = 0.57$.

3.

```{r}
2.08 + 1.96*0.35
2.08 - 1.96*0.35

-0.56 + 1.96*0.09
-0.56 - 1.96*0.09
```

In my model above, the relevant coefficients are $\gamma_2$ and $\beta_2$. A 95% confidence interval for $\gamma_2$ is [1.39, 2.77] and a 95% confidence interval for $\beta_2$ is [-0.75, -0.38].

4.

$P(Y_i \geq 1) = 1 - P(Y_i = 0) = 1 - (e^{-\lambda_i}(1 - p_i) + p_i)$

So we need to calculate $\widehat{p}_i$ and $\widehat{\lambda}_i$. We are told that $Persons = 3$, $Child = 0$, $Camper = 0$, and $LOS = 3$. So,

$\log \left( \dfrac{\widehat{p}_i}{1 - \widehat{p}_i} \right) = 1.81 - 0.98(3) = -1.13$

and thus $\widehat{p}_i = 0.244$. Similarly,

$\log (\widehat{\lambda}_i) = -1.13 + 0.45(3) + log(3) = 1.32$

and so $\widehat{\lambda}_i = 3.74$. The estimated probability of catching at least one fish is therefore $1 - (e^{-3.74}(1 - 0.244) + 0.244) = 0.74$.


5.

We are interested in testing the coefficient on $Child$ for the logistic part of the model. Using the model from question 2, our hypotheses are $H_0: \gamma_2 = 0$ vs. $H_A: \gamma_2 > 0$ (remember that the logistic component models the probability the group does *not* go fishing). We use a Wald test; the test statistic is 6.01, and the p-value is

```{r}
pnorm(6.01, lower.tail=F)
```


6.

We are interested in testing the coefficient on $Child$ in the Poisson part of the ZIP model. Using the model from question 2, our hypotheses are $H_0: \beta_2 = 0$ vs. $H_A: \beta_2 < 0$. We use a Wald test; the test statistic is -6.16, and the p-value is

```{r}
pnorm(-6.16, lower.tail=T)
```

