---
title: "Class activity solutions"
format: html
editor: source
author: "Ciaran Evans"
---

1. 

```{r}
set.seed(11)

rqpois <- function(n, mean, dispersion){
  return(rnbinom(n, mu = mean, size = mean/(dispersion - 1)))
}

n <- 200
nsim <- 1000

pearson_ests <- rep(NA, nsim)
mean_dev_ests <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n)
  y <- rqpois(n, mean = exp(3 + 0.1*x), dispersion = 5)
  
  m1 <- glm(y ~ x, family = poisson)
  pearson_resids <- residuals(m1, type = "pearson")
  pearson_ests[i] <- sum(pearson_resids^2)/m1$df.residual
  mean_dev_ests[i] <- m1$deviance/m1$df.residual
}

mean(pearson_ests)
mean(mean_dev_ests)


sd(pearson_ests)
sd(mean_dev_ests)
```

$\mathbb{E}[\widehat{\phi}] \approx 5$ for both estimates. The mean of the Pearson estimate is a little closer to the true dispersion parameter. The mean deviance estimate is slightly less variable.

2.

```{r}
qqplot(pearson_ests*m1$df.residual/5, rchisq(nsim, m1$df.residual),
       main = "Pearson estimate", 
       xlab = "Quantiles of Pearson estimates",
       ylab = "Quantiles of chi-square distribution")
abline(a=0, b=1)
```

```{r}
qqplot(mean_dev_ests*m1$df.residual/5, rchisq(nsim, m1$df.residual),
       main = "Mean deviance estimate", 
       xlab = "Quantiles of mean deviance estimates",
       ylab = "Quantiles of chi-square distribution")
abline(a=0, b=1)
```

The $\chi^2_{n-p}$ approximation is pretty good for both estimates. Maybe slightly better for the Pearson estimate.

3.

```{r}
n <- 200
nsim <- 1000

pearson_ests <- rep(NA, nsim)
mean_dev_ests <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n)
  y <- rqpois(n, mean = exp(0 + 0.1*x), dispersion = 5)
  
  m1 <- glm(y ~ x, family = poisson)
  pearson_resids <- residuals(m1, type = "pearson")
  pearson_ests[i] <- sum(pearson_resids^2)/m1$df.residual
  mean_dev_ests[i] <- m1$deviance/m1$df.residual
}

mean(pearson_ests)
mean(mean_dev_ests)


sd(pearson_ests)
sd(mean_dev_ests)
```

The Pearson estimate is still close to $\phi$ in expectation, but the mean deviance estimate is quite biased. The mean deviance estimate is less variable than the Pearson estimate.

4. 

```{r}
qqplot(pearson_ests*m1$df.residual/5, rchisq(nsim, m1$df.residual),
       main = "Pearson estimate", 
       xlab = "Quantiles of Pearson estimates",
       ylab = "Quantiles of chi-square distribution")
abline(a=0, b=1)
```

```{r}
qqplot(mean_dev_ests*m1$df.residual/5, rchisq(nsim, m1$df.residual),
       main = "Mean deviance estimate", 
       xlab = "Quantiles of mean deviance estimates",
       ylab = "Quantiles of chi-square distribution")
abline(a=0, b=1)
```

The $\chi^2_{n-p}$ approximation fails for both estimates. So, the Pearson estimate still does a good job estimating $\phi$, but it no longer has a $\chi^2_{n-p}$ distribution.