---
title: "Class activity solutions"
format: html
editor: source
author: "Ciaran Evans"
---

1. 

```{r}
n <- 100
nsim <- 1000
results <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n, sd = 0.5)
  y <- rpois(n, lambda = exp(x))
  
  m1 <- glm(y ~ x, family = poisson)
  
  upper <- summary(m1)$coefficients[2,1] + 
      1.96*summary(m1)$coefficients[2,2]
  lower <- summary(m1)$coefficients[2,1] - 
      1.96*summary(m1)$coefficients[2,2]
  
  results[i] <- upper > 1 && lower < 1
}

mean(results)
```

2. 

```{r}
n <- 100
r <- 10
nsim <- 1000
results <- rep(NA, nsim)

for(i in 1:nsim){
  x <- rnorm(n, sd = 0.5)
  y <- rnbinom(n, size = r, mu = exp(x))
  
  m1 <- glm(y ~ x, family = poisson)
  
  upper <- summary(m1)$coefficients[2,1] + 
      1.96*summary(m1)$coefficients[2,2]
  lower <- summary(m1)$coefficients[2,1] - 
      1.96*summary(m1)$coefficients[2,2]
  
  results[i] <- upper > 1 && lower < 1
}

mean(results)
```

Coverage is still reasonably close to 95%.

3.

```{r}
n <- 100
nsim <- 1000
rs <- c(0.5, 2, 5, 10, 20)
coverage <- rep(NA, length(rs))

for(j in 1:length(rs)){
  results <- rep(NA, nsim)

  for(i in 1:nsim){
    x <- rnorm(n, sd = 0.5)
    y <- rnbinom(n, size = rs[j], mu = exp(x))
    
    m1 <- glm(y ~ x, family = poisson)
    
    upper <- summary(m1)$coefficients[2,1] + 
        1.96*summary(m1)$coefficients[2,2]
    lower <- summary(m1)$coefficients[2,1] - 
        1.96*summary(m1)$coefficients[2,2]
    
    results[i] <- upper > 1 && lower < 1
  }
  
  coverage[j] <- mean(results)
}


plot(rs, coverage)
```

When the distribution is sufficiently non-Poisson (e.g., $r = 0.5$), coverage is much less than 95%. However, coverage can still be close to 95% when the distribution is close enough to a Poisson distribution, even though the Poisson assumption is wrong.


4.

```{r, message=F, warning=F}
library(statmod)
library(tidyverse)

n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x))

m1 <- glm(y1 ~ x, family = poisson)

data.frame(x = x, resids = qresid(m1)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "x", y = "Quantile residuals") 
```

When the Poisson regression assumptions are satisfied, the quantile residual plot looks like we would expect: the residuals are mostly between -2 and 2, they are randomly scattered around 0 with no clear pattern, and the variance seems constant.

5. 

```{r, warning=F, message=F}
n <- 1000
x <- rnorm(n, sd = 0.5)
y2 <- rnbinom(n, size=0.5, mu=exp(x))

m2 <- glm(y2 ~ x, family = poisson)

data.frame(x = x, resids = qresid(m2)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "x", y = "Quantile residuals") 
```

Breaking the Poisson distribution assumption, we can see that there is a lot more variability to the residuals, and the variance appears to change with x.

6. 

```{r, message=F, warning=F}
n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x^2))

m1 <- glm(y1 ~ x, family = poisson)

data.frame(x = x, resids = qresid(m1)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "x", y = "Quantile residuals") 
```

Breaking the shape assumption, there is a clear pattern to the quantile residual plot. The u-shape indicates that we may need to consider a quadratic function of x.

7. Changes in variance indicate potential violations of the distribution assumption. Patterns indicate potential violations of the shape assumption.
