---
title: "Class activity solutions"
format: html
editor: source
author: "Ciaran Evans"
---

1. 

```{r}
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")

m1 <- glm(Dengue ~ Age + WBC + PLT, data = dengue, family = binomial)
m1$deviance
```

The residual deviance is 5200.823

2. 

```{r}
-2*sum(m1$y * log(m1$fitted.values/(1 - m1$fitted.values)) + log(1 - m1$fitted.values))

```

It matches!

3. 

```{r}
y <- 0
p <- seq(0, 1, 0.01)
plot(p, y*log(p/(1 - p)) + log(1-p))
```

If $y = 0$, the maximum of $y \log \left( \frac{p}{1 - p} \right) + \log(1 - p)$ occurs at $p = 0$.

```{r}
y <- 1
p <- seq(0, 1, 0.01)
plot(p, y*log(p/(1 - p)) + log(1-p))
```

If $y = 1$, the maximum of $y \log \left( \frac{p}{1 - p} \right) + \log(1 - p)$ occurs at $p = 1$.

4. 

$$-2 \sum \limits_{i=1}^n \left( Y_i \log \left( \frac{\widehat{p}_i}{1 - \widehat{p}_i} \right) + \log(1 - \widehat{p}_i) \right)$$

The inner term $Y_i \log \left( \frac{\widehat{p}_i}{1 - \widehat{p}_i} \right) + \log(1 - \widehat{p}_i)$ is greater when $\widehat{p}_i$ is closer to $Y_i$ (see question 3). Since there is a negative sign out the front, the deviance is smaller when $\widehat{p}_i$ is closer to $Y_i$, so the deviance measures how close the predictions are (overall) to the observed data.

5. 

```{r}
y <- 5
lambda <- seq(1, 10, 0.1)
plot(lambda, y*log(lambda) - lambda)
```

The maximum occurs at $\lambda = 5$.

6. 

$t(y, \mu) = y \log \mu - \mu$, so $t(y, y) = y \log y - y$. Then,

$$d(y, \mu) = 2(y \log y - y - y \log \mu + \mu) = 2 \left( y \log \left( \frac{y}{\mu} \right) - (y - \mu) \right)$$

7. 

```{r}
y <- 5
mu <- seq(1, 10, 0.1)
d <- 2*(y*log(y/mu) - (y - mu))
plot(mu, d)

d[which.min(d)]
mu[which.min(d)]
```


8. 

```{r}
library(foreign)
articles <- read.dta("http://www.stata-press.com/data/lf2/couart2.dta")

m2 <- glm(art ~ phd + ment, data = articles, family = poisson)
m2$deviance
```


The residual deviance is 1669.352

9. 

```{r}
2*sum(m2$y * log(m2$y/m2$fitted.values), na.rm=T) - 2*sum(m2$y - m2$fitted.values)
```

10.

```{r}
2*(sum(log(1/factorial(m2$y))) + 
     sum(m2$y * log(m2$y) - m2$y,na.rm=T) - 
  (sum(log(1/factorial(m2$y))) + 
     sum(m2$y * log(m2$fitted.values) - m2$fitted.values)))
```

11. 

$$\log L = \sum \limits_{i=1}^n (Y_i \log \widehat{p}_i + (1 - Y_i) \log(1 - \widehat{p}_i))$$

so for the saturated model,

$$\log L_{saturated} = \sum \limits_{i=1}^n (Y_i \log Y_i + (1 - Y_i) \log(1 - Y_i)) = 0$$

