---
title: "Class activity solutions"
format: html
editor: source
author: "Ciaran Evans"
---

1.

```{r, message=F, warning=F}
library(statmod)
library(tidyverse)

n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x))

m1 <- glm(y1 ~ x, family = poisson)

data.frame(x = x, resids = qresid(m1)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "x", y = "Quantile residuals") 
```

When the Poisson regression assumptions are satisfied, the quantile residual plot looks like we would expect: the residuals are mostly between -2 and 2, they are randomly scattered around 0 with no clear pattern, and the variance seems constant.

2. 

```{r, warning=F, message=F}
n <- 1000
x <- rnorm(n, sd = 0.5)
y2 <- rnbinom(n, size=0.5, mu=exp(x))

m2 <- glm(y2 ~ x, family = poisson)

data.frame(x = x, resids = qresid(m2)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "x", y = "Quantile residuals") 
```

Breaking the Poisson distribution assumption, we can see that there is a lot more variability to the residuals, and the variance appears to change with x.

3. 

```{r, message=F, warning=F}
n <- 1000
x <- rnorm(n, sd = 0.5)
y1 <- rpois(n, lambda = exp(x^2))

m1 <- glm(y1 ~ x, family = poisson)

data.frame(x = x, resids = qresid(m1)) |>
  ggplot(aes(x = x, y = resids)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(x = "x", y = "Quantile residuals") 
```

Breaking the shape assumption, there is a clear pattern to the quantile residual plot. The u-shape indicates that we may need to consider a quadratic function of x.

4. Changes in variance indicate potential violations of the distribution assumption. Patterns indicate potential violations of the shape assumption.
